Open Clusters Characterization in Gaia DR2 Using ML Algorithms

### 2020-11-23 thesis/thesis.tex
thesis: Add more content to Method chapter

\frontmatter
\tableofcontents
\listoffigures
\listoftables

@@

Stars that belong to a same OC share relative positions, so their parallax value is similar among them. They also share similar values of
proper motion, both in right ascension and declination. Another property, since these stars were born from the same gas cloud and at the same
temporary stage, their metallicity must be uniform. However, to take into account this last property, we face the inconvinient that this
parameter is not directly available in Gaia DR2 database. We will avoid this issue by looking at $E(b-r) / G_{mag}$ diagrams with well
defined isochrone curves \cite{bressan2012parsec}. These isochrone curves are derived from theoretical models that mainly involve metallicity
and mass/brightness from stars. With these models we can define important properties for the development of stars and their evolution in the
H-R diagram. Examples of these curves are shown in Figure \ref{fig:examples_of_isochrones}.

@@

To achieve the general goal, we'll follow the following milestones:

@@

@@

% TODO: Hablar sobre los algoritmos de clustering que se van a emplear: K-Means, DEC

@@

In this chapter we address the different steps taken to perform the creation of the unsupervised clustering model for open cluster characterization.

All code is developed with the Python programming language \cite{Python3}. Other auxiliary tools such as \emph{Docker} \cite{merkel2014docker},
\emph{PostgreSQL} \cite{postgresql}, \emph{Jupyter Notebooks} \cite{Kluyver2016jupyter}, and \emph{Astropy} \cite{astropy:2013} \cite{astropy:2018},
\emph{Scikit-Learn} \cite{scikit-learn}, \emph{seaborn} \cite{michael_waskom_2017_883859}, \emph{SQLAlchemy} \cite{sqlalchemy} and
\emph{Keras} \cite{chollet2015keras} frameworks have been used too.

As mentioned before, we want to characterize clusters looking at their dynamics properties.
So, these are the main features we will take into account for clustering:

\begin{itemize}
  \item OCs and stars \emph{right ascension}: $\alpha$ and \emph{declination}: $\delta$
  \item OCs \emph{estimated sizes}: $r$
  \item \emph{Proper motion}: $\mu_{\alpha}$, $\mu_{\delta}$ and \emph{parallax}: $\varpi$
  \item \emph{Associated errors} will be taken into account to estimate prediction errors
\end{itemize}

@@

As mentioned before, I have chosen the Gaia DR2 database as the data source for this work. Thus, the first step is to download the required data from Gaia DB
and store it in a custom database for later access.

Due to the large amount of data inside Gaia database, a complete download is not viable neither useful.
In order to reduce the size of the dataset to be downloaded, the OpenClust catalogue \cite{dias2002new} (Figure \ref{fig:OpenClustComplete})
has been used to make a preselection of sky regions.

This download is not limited to the size registered in the catalogue. Instead, a wider region is downloaded for each cluster to include
stars outside the cluster and take them into account when training the model so the model will be able to take apart those outsiders
when characterizing the cluster.

@@

The downloaded dataset covers nearly 114 million stars (42GB of compressed data), that is significantly smaller than the whole Gaia DR2 dataset, which contains
information for approximately 1,600 million stars, but it is still being too large.

Since not all the downloaded clusters are good enough, the following filters are applied to discard some clusters which don't have enough stars or contain too
many \emph{null} values:

\begin{itemize}
  \item Cluster diameter above 25.0 arcmin
  \item Parallax absolute value greater than 0.0
  \item Number of stars in the selected region above 40,000 stars
\end{itemize}

As shown in Figure \ref{fig:OpenClustSelection} this selection stills being a good representation of all clusters in the Milky Way since they are
equally distributed around the galaxy line.

After this selection, the number of cluster to look up is 169 with nearly 75 million stars.

Since there is no upper limit to the number of stars inside a cluster, for the sake of simplicity and as a commitment to the project delivery dates
and the available computing power, smaller clusters will be preferred over the greatest ones.

As an extension of this work, the designed model could be applied to a region not covered by this preselection for a later understanding of the results.

@@

@@

The first one is a container built from a \verb|python:3.8| image that contains the \verb|cdalvaro| package and the \verb|downloader.py| script.
This script is prepared to load the OpenClust catalogue, connect to the Gaia DR2 database, download all stars contained inside each cluster
(with a radius multiplier to increase the area to be downloaded) and save the downloaded data inside the custom database hosted in the second container.

Downloader image build instructions are available at \href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/downloader.Dockerfile}{downloader.Dockerfile}
and latest image is available for pulling from \href{https://github.blog/2020-09-01-introducing-github-container-registry/}{GitHub Container Registry}:

@@

This image is automatically built with \href{https://github.com/features/actions}{GitHub Actions} on every push to the \verb|main| branch.

@@

The database has two main tables, \verb|regions| for storing cluster properties such as location, diameter and other properties, and \verb|gaiadr2_source|
which contains data for all downloaded stars. This second table is partitioned by region, storing all stars related to one cluster inside an independent
table for faster access.

@@

These containers are orchestrated with a \href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/docker-compose.yml}{docker-compose.yml}
file which automatically launches one instance of each container.

The code developed for this work can be found inside the \verb|src| directory at GitHub:
\href{https://github.com/cdalvaro/machine-learning-master-thesis}{cdalvaro/machine-learning-master-thesis}

@@

@@

@@

>>>>

\frontmatter
\tableofcontents

@@

Stars that belong to a same OC share relative positions, so their parallax value is similar among them. They also share similar values of
proper motion, both in right ascension and declination. Another property, since these stars were born from the same gas cloud and at the same
temporary stage, their metallicity must be uniform. However, to take into account this last property, we face the inconvenient that this
parameter is not directly available in Gaia DR2 database. We will avoid this issue by looking at $E(b-r) / G_{mag}$ diagrams with well
defined isochrone curves \cite{bressan2012parsec}. These isochrone curves are derived from theoretical models that mainly involve metallicity
and mass/brightness from stars. With these models we can define important properties for the development of stars and their evolution in the
Hertzsprung–Russell (H-R) diagram. Examples of these curves are shown in Figure \ref{fig:examples_of_isochrones}.

@@

To achieve the general goal, we will follow the following milestones:

@@

% TODO: Hablar de la misión Gaia
% TODO: Hablar de KMedias y DEC
% TODO: Mencionar trabajos relacionados

@@

@@

In this chapter we address the different steps followed to perform the creation of the unsupervised clustering model for open cluster characterization.

All code has been developed with the Python programming language \cite{Python3}. Other auxiliary tools like \emph{Docker} \cite{merkel2014docker},
\emph{PostgreSQL} \cite{postgresql}, \emph{Jupyter Notebooks} \cite{Kluyver2016jupyter}, and frameworks such as \emph{Astropy} \cite{astropy:2013} \cite{astropy:2018},
\emph{Scikit-Learn} \cite{scikit-learn}, \emph{Seaborn} \cite{michael_waskom_2017_883859}, \emph{SQLAlchemy} \cite{sqlalchemy} and
\emph{Keras} \cite{chollet2015keras} have been used too.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/raw_pm_melotte_22.pdf}
  \end{subfigure}
  \caption{Melotte 22 proper motions}
  \label{fig:raw_pm_melotte_22}
\end{figure}

For the sake of simplicity, we are taking \emph{Melotte 22 (Messier 45)} \cite{elsanhoury2019ppmxl} to illustrate the whole method. This cluster is well studied
and their stars are not too mixed with other region stars.

Figure \ref{fig:raw_pm_melotte_22} shows \emph{proper motions in right ascension and declination} for a sample of the downloaded dataset for Melotte 22.
At first sight, two main clusters can be distinguished, one of them centered nearly at (0, 0) and the second cluster with center at (20, -45). This second cluster
is the one we are looking for.

However, although this second cluster is almost isolated, there are stars that do not belong to the OC. Thus, we need more information to properly characterize the OC.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/raw_parallax_melotte_22.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/raw_hr_diagram_melotte_22.pdf}
    \end{subfigure}
  \end{subfigure}
  \caption{Melotte 22 parallax and H-R diagram}
  \label{fig:raw_parallax_hr_diagram_mellote_22}
\end{figure}

Figure \ref{fig:raw_parallax_hr_diagram_mellote_22} shows the parallax histogram and the H-R diagram for this cluster, respectively.

The first one shows a resonance at $\approx 7.5mas$ belonging the OC. While the second figure would help us to look for isochrone curves and so,
to determine the age of these stars.

@@

Before being able to develop and test our clustering model, the first step is to download the required data from Gaia repository
and store it in a custom database for later access.

Due to the large amount of data available at Gaia, a complete download is not viable neither useful.
In order to reduce the size of the dataset to be downloaded, the OpenClust catalogue \cite{dias2002new} (Figure \ref{fig:OpenClustComplete})
has been used to restrict the sky regions to be explored.

This download is not limited to the size registered in the catalogue. Instead, a wider region is downloaded for each cluster to include
stars outside the cluster. The idea is that the unsupervised model must be able to clusterize the stars that belongs to the OC and to discard
those outsiders when characterizing the open cluster.

@@

Taking these considerations into account, the downloaded dataset covers nearly 114 million stars (42GB of compressed data). This dataset is significantly
smaller than the whole Gaia DR2 dataset, which contains information for approximately 1,600 million stars. However, it is still being too large.

Since not all the downloaded clusters are good enough, we will apply a series of filters to discard clusters which do not have enough stars or contain too
many \emph{null} values.

So, a cluster must fulfil the following filters to be accepted:

\begin{itemize}
  \item Cluster diameter above 25.0 arcmin
  \item Parallax absolute value greater than 0.0
  \item Number of stars\footnote{Every star must have all required features completely defined, i.e. without null values} in the selected region above 40,000 stars
\end{itemize}

As shown in Figure \ref{fig:OpenClustSelection}, these constrains give us a smaller dataset but it is still being a good representation of all clusters in the Milky Way
since they are equally distributed around the galaxy disk.

After having applied these filters, the number of cluster to analyse is 169 with nearly 75 million stars.

Since we have set no upper limit to the number of stars inside a cluster, for the sake of simplicity and as a commitment to the project delivery dates
and the available computing power, smaller clusters will be preferred over the greatest ones.

As an extension of this work, the designed model could be applied to a region not covered by these constrains for a later understanding of the results.

@@

\subsection{Download Process}

@@

The first one is a container built from a \verb|python:3.8| image that contains the \verb|cdalvaro| package and the \verb|downloader.py| script.
This script is prepared to load the OpenClust catalogue, connect to the Gaia DR2 database, download all stars contained inside each cluster
(with a radius multiplier to increase the area to be downloaded) and save the downloaded data inside our custom database hosted in the second container.

\verb|downloader.Dockerfile| contains the build instructions for the downloader image. We use \href{https://github.com/features/actions}{GitHub Actions}
for automating the process of building a new version of the image with every new push made to the \verb|main| branch.

This image can be pulled from the \href{https://github.blog/2020-09-01-introducing-github-container-registry/}{GitHub Container Registry}:

@@

@@

The database has two main tables in \verb|public| schema:

\begin{itemize}
  \item \verb|regions| for storing cluster properties such as \emph{location}, \emph{diameter} and other properties
  \item \verb|gaiadr2_source| which contains data for all downloaded stars from Gaia
\end{itemize}

The second table is partitioned by region, storing all stars related to one cluster inside an independent table for optimized access performance.

@@

These containers are orchestrated with a \verb|docker-compose.yml| file which automatically launches one instance of each container.

The developed code for this work can be found inside the \verb|src| directory at GitHub:
\href{https://github.com/cdalvaro/machine-learning-master-thesis}{cdalvaro/machine-learning-master-thesis} repository.

@@

% TODO: Hablar sobre la selección de variables

@@

As mentioned before, we want to characterize clusters by looking at their dynamics properties.
So, these are the main features we will take into account for clustering:

\begin{itemize}
  \item OCs and stars \emph{right ascension}: $\alpha$ and \emph{declination}: $\delta$
  \item OCs \emph{estimated sizes}: $r$
  \item \emph{Proper motion}: $\mu_{\alpha}$, $\mu_{\delta}$ and \emph{parallax}: $\varpi$
  \item \emph{Associated errors} will be taken into account to discard data with excessive uncertainty
\end{itemize}

@@

\listoffigures
\listoftables
